---
title: "Linearity_Normality"
output: pdf_document
date: "2024-11-20"
---


```{r}
ggplot(data, aes(x = age)) +
  geom_histogram(binwidth = 5, fill = "skyblue", color = "black") +
  theme_minimal() +
  labs(title = "Age Distribution", x = "Age", y = "Frequency")

```

```{r}
ggplot(data, aes(x = creatinine_phosphokinase)) +
  geom_histogram(binwidth = 100, fill = "coral", color = "black") +
  theme_minimal() +
  labs(title = "Distribution of CPK Levels", x = "CPK Level", y = "Frequency")

```
```{r}
numeric_data <- data %>% select_if(is.numeric)
cor_matrix <- cor(numeric_data)

library(corrplot)
corrplot(cor_matrix, method = "circle", type = "upper", tl.col = "black", tl.cex = 0.8)

```
```{r}
log_model <- glm(DEATH_EVENT ~ age + ejection_fraction + serum_creatinine + serum_sodium, 
                 data = data, family = "binomial")

summary(log_model)

exp(cbind(OR = coef(log_model), confint(log_model)))

```


```{r}
library(GGally)

ggpairs(data, aes(color = as.factor(DEATH_EVENT), alpha = 0.7),
        columns = c("age", "ejection_fraction", "serum_creatinine", "serum_sodium"))

```

```{r}
library(rpart)
library(rpart.plot)

tree_model <- rpart(DEATH_EVENT ~ ., data = data, method = "class")

rpart.plot(tree_model, type = 4, extra = 104, cex = 0.8, main = "Decision Tree")

```

```{r}
data <- read.csv("heart_failure_clinical_records_dataset.csv")

continuous_vars <- c("age", "creatinine_phosphokinase", "ejection_fraction", 
                     "platelets", "serum_creatinine", "serum_sodium")

par(mfrow = c(2, 2)) # Set up a grid for multiple plots
for (var in continuous_vars) {
  
  hist(data[[var]], main = paste("Histogram of", var), xlab = var, col = "skyblue", border = "white")
  
  qqnorm(data[[var]], main = paste("Q-Q Plot of", var))
  qqline(data[[var]], col = "red")
}

for (var in continuous_vars) {
  test <- shapiro.test(data[[var]])
  cat("Shapiro-Wilk test for", var, ":\n")
  print(test)
}
```



```{r}
model <- glm(DEATH_EVENT ~ age + serum_creatinine + ejection_fraction, data=data, family=binomial)

fitted_values <- fitted(model)
residuals <- residuals(model, type="pearson")
plot(fitted_values, residuals, main="Residuals vs Fitted Values", xlab="Fitted Values", ylab="Residuals")
abline(h=0, col="red")

```

```{r}
par(mfrow = c(2, 2)) 
for (var in continuous_vars) {
  plot(data[[var]], data$DEATH_EVENT, 
       main = paste(var, "vs DEATH_EVENT"), 
       xlab = var, ylab = "DEATH_EVENT", pch = 19, col = "blue")
  abline(lm(data$DEATH_EVENT ~ data[[var]]), col = "red")
}


for (var in continuous_vars) {
  model <- glm(DEATH_EVENT ~ data[[var]], data = data, family = binomial)
  summary(model)
}


model_quad <- glm(DEATH_EVENT ~ data[[var]] + I(data[[var]]^2), family = binomial, data = data)
summary(model_quad)

```


```{r}
data$log_odds <- predict(model, type = "link")  # 'link' gives log-odds

continuous_vars <- c("age", "serum_creatinine", "ejection_fraction", 
                     "creatinine_phosphokinase", "platelets", "serum_sodium", "time")

par(mfrow = c(3, 2)) # Set grid for multiple plots
for (var in continuous_vars) {
  plot(data[[var]], data$log_odds,
       main = paste(var, "vs Log-Odds"),
       xlab = var, ylab = "Log-Odds", col = "blue", pch = 19)
  abline(lm(data$log_odds ~ data[[var]]), col = "red")  # Add a linear trendline
}


poly_model <- glm(DEATH_EVENT ~ poly(age, 2) + poly(serum_creatinine, 2) +
                  poly(ejection_fraction, 2) + poly(creatinine_phosphokinase, 2) +
                  poly(platelets, 2) + poly(serum_sodium, 2) + poly(time, 2) +
                  anaemia + diabetes + high_blood_pressure + sex + smoking,
                  family = binomial, data = data)

summary(poly_model)

```

## For all variables tested, the Shapiro-Wilk test produced p-values below 0.05, meaning none of these variables follow a normal distribution

## almost all of the histograms were either skewed right or left which indicates a low level of normality. the platlets and serum sodium were the ones the were the most normally dsitributed. 

## using the qq plots i can see that age, ejection, serum sodium, and platlets seem to be the most normal. the other variables were not as normal. 

## The logistic regression results show that the predictor variables (like age or serum creatinine) are significantly linked to the likelihood of a DEATH_EVENT. The model seems to fit well, as shown by the lower residual deviance. 

## adding the squared term didn't help the model much. This means the relationship between the variables (like age or creatinine levels) and the risk of death looks pretty straight-line, not curvy. So, keeping it simple with a linear model makes the most sense here. 

