---
title: "Outline"
author: "Paige Galvan, Neha Deshpande, and Witlie Leslie"
output: pdf_document
date: "2024-11-12"
---

## Questions/Big Ideas:

Does location/gender/age affect the after health recovery?



```{r}
# Data 

data <- 'Esophageal_Dataset.csv'
data <- read.csv(data)
summary(data)

```

## 1st Step 
Go through play w the data 
clean data 


Variables we like: 
main - cancer outcome: primary_pathology_karnofsky_performance_score & primary_pathology_eastern_cancer_oncology_group


three subsets:
- demographics (gender, height, weight, country origin, race)
- lifestyle
- medical

## Demographics 
```{r}

demogrpahics_data <- dplyr::select(data,
                                   gender,
                                   country_of_birth,
                                   race_list,
                                   height, 
                                   weight)
demogrpahics_data <- na.omit(demogrpahics_data)

cor_matrix <- cor(demogrpahics_data[, -7], use = "complete.obs")
cor_matrix

```

## Medical
```{r}
medical_data <- data[, c("icd_o_3_site", "icd_o_3_histology", "icd_10", "vital_status", "days_to_death")]r
```


## Analysis Types 

Lasso 
CV

```{r}

Esophageal_Dataset <- read_csv("Esophageal_Dataset.csv")

lifestyle_data <- dplyr::select(Esophageal_Dataset,
                                age_began_smoking_in_years, 
                                number_pack_years_smoked,
                                frequency_of_alcohol_consumption, 
                                amount_of_alcohol_consumption_per_day,
                                height, 
                                weight, 
                                primary_pathology_age_at_initial_pathologic_diagnosis)

lifestyle_data <- na.omit(lifestyle_data)
cor_matrix <- cor(lifestyle_data[, -7], use = "complete.obs")
cor_matrix

corrplot(cor_matrix, method = "circle", type = "upper", tl.col = "black", tl.srt = 45)

ggplot(lifestyle_data, aes(x = as.factor(frequency_of_alcohol_consumption))) +
  geom_bar(fill = "skyblue", color = "black", alpha = 0.7) +
  labs(title = "Frequency of Alcohol Consumption", x = "Frequency of Alcohol Consumption", y = "Count") +
  theme_minimal()

ggplot(lifestyle_data, aes(x = as.factor(age_began_smoking_in_years))) +
  geom_bar(fill = "lightgreen", color = "black", alpha = 0.7) +
  labs(title = "Age Began Smoking", x = "Age Began Smoking (Years)", y = "Count") +
  theme_minimal()

ggplot(na.omit(Esophageal_Dataset[, c("alcohol_history_documented", "person_neoplasm_cancer_status")]), 
       aes(x = alcohol_history_documented, fill = person_neoplasm_cancer_status)) +
  geom_bar(position = "fill") +
  labs(title = "Alcohol Consumption and Esophageal Cancer Status", x = "Alcohol Consumption History", y = "Proportion", fill = "Cancer Status") +
  theme_minimal()



```


hi



