---
title: "Predicting Heart Failure from Risk Factor Data"
author: "Paige Galvan, Neha Deshpande, & Witlie Leslie"
date: "2024-11-25"
output:
  beamer_presentation: default
  powerpoint_presentation: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```


```{r, include = FALSE}
library(caTools) 
library(randomForest) 
library(e1071) 
library(ggplot2)
library(dplyr)
library(tidyr)
library(readr)
library(MASS)
library(GGally)
library(tidymodels)
library(ISLR)
library(class)
library(reshape2)
library(pROC)
```

## Introduction (Witlie)

For our project, we aimed to create a model that can predict the event of death as a result of heart failure from clinical patient data.

- Binary response variable "DEATH_EVENT", value of 1 indicating patient deceased (0 otherwise)
- 12 risk factor variables
    - 5 binary: anemia status, diabetes status, high blood pressure status, sex, and smoking status
    - 7 numerical: age, creatine phosphokinase level, ejection fraction, platelet concentration, serum creatine level, serum sodium level, and length of follow-up period

# Motivation (Neha)

- Investigating heart failure, a condition impacting millions worldwide, and exploring its complex causes and contributing factors.

- Spotting patterns between risk factors and how heart failure progresses to get a better understanding.


# Linearity and Normality (Neha)

Normality

- Objective: Assess whether continuous predictor variables are normally distributed.

- Variables analyzed:
Age, Creatinine Phosphokinase, Ejection Fraction, Platelets, Serum Creatinine, Serum Sodium.


- Testing residuals ensures the model fits the data well and detects patterns that might indicate a need for model improvement.

- Most variables were either right or left skewed, showing low normality, except for platelets and serum sodium, which were closer to normal

- Tests performed: 

Shapiro-Wilk Test:
p-values < 0.05 for all variables → None follow a normal distribution.


# Graphs
```{r, echo=FALSE, warning=FALSE, message=FALSE}
data <- read.csv("heart_failure_clinical_records_dataset.csv")

# Select a few continuous variables
selected_vars <- c("age", "serum_creatinine")

# Set up the plotting area to display two histograms and two Q-Q plots on one slide
par(mfrow = c(2, 2)) # 2 rows and 2 columns

# Loop through the selected variables to create histograms and Q-Q plots
for (var in selected_vars) {
  
  # Histogram for the selected variable
  hist(data[[var]], main = paste("Histogram of", var), 
       xlab = var, col = "skyblue", border = "white")
  
  # Q-Q Plot for the selected variable
  qqnorm(data[[var]], main = paste("Q-Q Plot of", var))
  qqline(data[[var]], col = "red")
}


```


# Linearity 

- Linearity assumes a straight-line relationship between the predictor variables and the log odds of the outcome in logistic regression

- Fitted linear trendlines show non linear relationships for most variables.

- Logistic regression results:

The residuals exhibit a curved pattern, indicating that the relationship between the predictors and the response variable may not be linear.


# Graphs 
```{r, echo=FALSE, warning=FALSE, message=FALSE}
model <- glm(DEATH_EVENT ~ age + serum_creatinine + ejection_fraction, data=data, family=binomial)

fitted_values <- fitted(model)
residuals <- residuals(model, type="pearson")
plot(fitted_values, residuals, main="Residuals vs Fitted Values", xlab="Fitted Values", ylab="Residuals")
abline(h=0, col="red")

```

# Polynomial Model to Test Linearity

- Polynomial models allow us to model nonlinear relationships between predictors and the outcome

-  The polynomial model identified several significant predictors of the likelihood of a death event, including age, serum creatinine, ejection fraction, and time. However, many other predictors, such as platelets, serum sodium, and factors like anaemia, diabetes, and smoking, were found to be statistically insignificant

# SVM & Random Forest (Paige)

# Logistic Regression (Witlie)

Logistic regression was an obvious first choice in tackling this binary classification problem. I created 3 models to assess predictive performance with different regularization methods.

- Ridge Regression
- Lasso
- No regularization

# Ridge Regression

```{r, echo=FALSE, warning=FALSE, message=FALSE}
set.seed(445)

library(tidymodels)
library(tidyverse)
library(ggplot2)
library(dplyr)

data <- read.csv("heart_failure_clinical_records_dataset.csv")

# scale continuous variables
numeric_vars <- sapply(data, is.numeric)
boolean_vars <- sapply(data, function(x) all(x %in% c(0,1)))

need_scale <- numeric_vars & !boolean_vars

data[need_scale] <- scale(data[need_scale])

# convert DEATH_EVENT to a factor 
data <- data %>%
  mutate(DEATH_EVENT = factor(DEATH_EVENT, levels = c("0", "1")))

data_split <- initial_split(data, prop = 0.8)
train_data <- training(data_split)
test_data <- testing(data_split)

prep_data <- recipe(DEATH_EVENT ~ ., data = train_data)

# create vector of 100 lambda values from 0.01 to 10^10
lambda <- lambda <- 10^seq(-2, 10, length.out = 100)
tune_df <- data.frame(lambda = lambda)

# fit ridge regression model for each lambda
prep_data <- recipe(DEATH_EVENT ~ ., data = data)

ridge_ests <- data.frame()

for(lam in lambda) {
  ridge_spec <- logistic_reg(mixture = 0, penalty = lam) |>
    set_mode("classification") |>
    set_engine("glmnet")
  
    workflow() |>
    add_model(ridge_spec) |>
    add_recipe(prep_data) |>
    fit(data) |>
    tidy() |>
    bind_rows(ridge_ests) -> ridge_ests
}

# visualize estimated coefficients for each lambda
ridge_ests |>
  filter(term != "(Intercept)") |>
  ggplot() +
  labs(title = "Ridge Regression: Coefficient Estimates vs Penalty") +
  geom_line(aes(penalty, estimate, group = term, colour = term)) +
  coord_trans(x = "log10")
```

# Ridge Regression

Using 10-fold cross-validation, I found the lambda with the highest ROC-AUC value of 0.887 was lambda = 0.0933

```{r, echo=FALSE, warning=FALSE, message=FALSE}
# perform 10-fold cross validation, estimate roc_auc for each lambda
data_10foldcv <- vfold_cv(data, v = 10)

ridge_spec <- logistic_reg(mixture = 0, penalty = tune("lambda")) |>
  set_mode("classification") |>
  set_engine("glmnet")

workflow() |>
  add_model(ridge_spec) |>
  add_recipe(prep_data) |>
  tune_grid(resamples = data_10foldcv, grid = tune_df, metrics = metric_set(roc_auc, accuracy)) -> ridge_tune

# determine best lambda

ridge_metrics <- ridge_tune |>
  collect_metrics() |>
  filter(.metric == "roc_auc") |>
  mutate(method = "Ridge")

ridge_best <- show_best(ridge_tune, metric = "roc_auc", n = 1)

ridge_tune |>
  collect_metrics() |>
  dplyr::select(lambda, .metric, mean) |>
  pivot_wider(names_from = .metric, values_from = mean) |>
  ggplot() +
  labs(title = "Ridge Regression: roc_auc vs lambda") +
  geom_line(aes(lambda, roc_auc)) + 
  geom_point(aes(lambda, roc_auc)) +
  geom_point(aes(ridge_best$lambda, ridge_best$mean), color = "red", shape = 1, size = 4) +
  coord_trans(x = "log10")
```

# Lasso

Lasso shows a steeper dropoff of coefficient estimates as a result of feature reduction

```{r, echo=FALSE, warning=FALSE, message=FALSE}
# fit lasso model for each lambda
lasso_ests <- data.frame()
for(lam in lambda) {
  lasso_spec <- logistic_reg(mixture = 1, penalty = lam) |>
    set_mode("classification") |>
    set_engine("glmnet")
  
    workflow() |>
    add_model(lasso_spec) |>
    add_recipe(prep_data) |>
    fit(data) |>
    tidy() |>
    bind_rows(lasso_ests) -> lasso_ests
}

# visualize estimated coefficients for each lambda
lasso_ests |>
  filter(term != "(Intercept)") |>
  ggplot() +
  labs(title = "Lasso Method: Coefficient Estimates vs Penalty") +
  geom_line(aes(penalty, estimate, group = term, colour = term)) +
  coord_trans(x = "log10")
```

# Lasso

Using 10-fold cross validation, I found the lambda with the highest ROC-AUC value of 0.886 was lambda = 0.01

```{r, echo=FALSE, warning=FALSE, message=FALSE}
# perform 10-fold cross validation, estimate roc_auc for each lambda
lasso_spec <- logistic_reg(mixture = 1, penalty = tune("lambda")) |>
  set_mode("classification") |>
  set_engine("glmnet")

workflow() |>
  add_model(lasso_spec) |>
  add_recipe(prep_data) |>
  tune_grid(resamples = data_10foldcv, grid = tune_df, metrics = metric_set(roc_auc, accuracy)) -> lasso_tune

lasso_metrics <- lasso_tune |>
  collect_metrics() |>
  filter(.metric == "roc_auc") |>
  mutate(method = "Lasso")

lasso_best <- show_best(lasso_tune, metric = "roc_auc", n = 1)

# show best lambda with lasso
lasso_tune |>
  collect_metrics() |>
  dplyr::select(lambda, .metric, mean) |>
  pivot_wider(names_from = .metric, values_from = mean) |>
  ggplot() +
  geom_line(aes(lambda, roc_auc)) +
  labs(title = "Lasso Method: ROC-AUC vs Lambda") +
  geom_point(aes(lambda, roc_auc)) +
  geom_point(aes(lasso_best$lambda, lasso_best$mean), color = "red", shape = 1, size = 4) +
  coord_trans(x = "log10")
```

# Results

- No Regularization:

- Ridge Regression:

- Lasso:


# SVM (Paige)

- Predicting Death Event 
- Like our Linear Regression methods we began by splitting the data into training and test data. 
- I began by assessing the support variables and determining the acurracy of the model. 


```{r, echo=FALSE, warning=FALSE, message=FALSE}
set.seed(42)
scaled_data <- data
library(e1071)             
split <- sample.split(scaled_data$DEATH_EVENT, SplitRatio = 0.8)
train_data <- subset(scaled_data, split == TRUE)
test_data <- subset(scaled_data, split == FALSE)

svm_model <- svm(DEATH_EVENT ~ ., data = train_data, kernel = "radial", cost = 1, scale = TRUE)


summary(svm_model)

predictions <- predict(svm_model, test_data)

conf_matrix <- table(Predicted = predictions, Actual = test_data$DEATH_EVENT)
print(conf_matrix)

accuracy <- sum(predictions == test_data$DEATH_EVENT) / length(predictions)
print(paste("Accuracy: ", round(accuracy, 3)))

```

- Accuracy is 81.7% which details our predictive model should be fairly sound. 



# SVM Model

```{r}
svm_data <- scaled_data[, c("age", "serum_creatinine", "DEATH_EVENT")]

svm_model_2d <- svm(DEATH_EVENT ~ age + serum_creatinine, data = svm_data, kernel = "radial")

x_range <- seq(min(svm_data$age) - 1, max(svm_data$age) + 1, length.out = 100)
y_range <- seq(min(svm_data$serum_creatinine) - 1, max(svm_data$serum_creatinine) + 1, length.out = 100)
grid <- expand.grid(age = x_range, serum_creatinine = y_range)

predictions_grid <- predict(svm_model_2d, grid)

ggplot(svm_data, aes(x = age, y = serum_creatinine)) +
  geom_point(alpha = 0.7, size = 3) +
  geom_tile(data = grid, 
            aes(x = age, y = serum_creatinine, 
                fill = predictions_grid), 
            alpha = 0.3) +
  scale_fill_manual(values = c("red", "blue")) +
  scale_color_manual(values = c("red", "blue")) +
  theme_minimal() +
  labs(title = "SVM Decision Boundary", x = "Age", y = "Serum Creatinine") +
  theme(legend.position = "none")

```

- The SVM Boundry demonstrates that 





# Random Forest (paige)



# Results (Paige)

# References 

Davide Chicco, Giuseppe Jurman: Machine learning can predict survival of patients with heart failure from serum creatinine and ejection fraction alone. BMC Medical Informatics and Decision Making 20, 16 (2020). https://bmcmedinformdecismak.biomedcentral.com/articles/10.1186/s12911-020-1023-5 
